### 2012-11-28 branches/CUDA/BookGPU/Chapters/chapter14/ch14.tex
AG updates to gpu chapter
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1822 40876441-3496-0410-a613-a2fca35b2ae2

In the CPU version, the \textit{Ludwig} implementation stores one
time level of distribution values $f_i(\mathbf{r}; t)$. This distribution
is stored as a flat array of C data type \texttt{double}, and laid out so
that all elements of the velocity are contiguous at a given site (often
referred to as ``array-of-structures'' order). This is appropriate
for sums over
the distribution required locally at the collision stage as illustrated
schematically in Listing~\ref{ch14:listing1}. (The temporary scalar
\texttt{a\_tmp} allows caching of the intermediate accumulated value
in the innermost loop.)
A variety of standard sequential optimisations are relevant for the
collision stage (loop unrolling, inclusion of SIMD operations, and
so on \cite{wellein2006}). 
For example, the collision stage in \textit{Ludwig} includes explicit
SIMD code, which is useful if the compiler on a given platform cannot
identify it.
The propagation stage is separate, and is organised as a ``pull''
(again, various optimisation have been considered, e.g., 
\cite{pohl2003,mattila2007,wittmann2012}).
No further optimisation is done
here beyond ensuring that the ordering of the discrete velocities allows
memory access to be as efficient as possible. While these optimisations
are important, it should be remembered that for some complex fluid
problems, the hydrodynamics embodied in the LB calculation is a relatively
small part of the total computational cost (at the 10\%-level in some cases).
This means optimisation effort may be better concentrated elsewhere.

@@

\begin{figure}[!t]
\centering
\includegraphics[width=12cm]{Chapters/chapter14/figures/decomphalo}
\caption{Left: the lattice is decomposed between MPI tasks. For clarity
we show a 2D decomposition of a 3D lattice, but in practice we decompose
in all 3 dimensions. Halo cells are added to each sub-domain (as shown on
the right for a single slice) which store data retrieved from remote
neighbours in the halo exchange. CAN WE ADD: Lower right: the D3Q19
velocity set required 5 ``outgoing'' elements to be transferred at
each edge.}
\label{ch14:fig:decomphalo}
\end{figure}

@@

In this section we describe the steps taken to enable \textit{Ludwig}
for the GPU. There are a number of crucial issues: first, the
minimisation of data traffic between host and device; second, the
maximisation of available parallelism and third, the issue of
memory coalescing. We discuss each of these in turn.

While the most important section of the code in terms of floating-point
performance is the collision stage, this cannot be the only consideration
for a GPU implementation. It is essential to offload all computational
activity which involves the main data structures (such as the distribution)
to the GPU. This includes kernels with relatively low computational demand,
such as the propagation stage. All relevant data then remain resident on
the GPU, which avoids expensive host-device data transfer at each iteration
of the algorithm. Such transfers would negate any benefit of GPU acceleration.
We note that for a complex fluid code, this requirement can extend to a
considerable number of kernels, although we limit the discussion to
collision and propagation for brevity.

@@

To demonstrate the effectiveness of this approach, we compare the
performance of both CPU and GPU versions of \textit{Ludwig}. To
test the complex fluid nature of the code, the problem is
actually an immiscible fluid mixture which uses a second distribution
function to introduce a composition variable. The interested reader
is referred to \cite{ch14:stratford-jsp2005} for further details. The
largest total problem size used is $1548\times 1764\times 1568$ ?PLEASE CHECK EXACTLY.
The CPU system features a Cray XE6 architecture with 2 16-core AMD Opteron
CPUs per node, and with nodes interconnected using Cray Gemini technology.
For GPU results, a Cray XK6 system is used: this is very similar to the
XE6, but has one CPU per node replaced with an NVIDIA X2090 GPU. Each
node in the GPU system therefore features a single Opteron CPU acting
as a host to a single GPU. The inter-node interconnect architecture is
the same as for the Cray XE6. The GPU performance tests use a prototype
Cray XK6 system with 936 nodes (the LARGEST AVAILABLE AT TIME OF WRITING?).
To provide a fair comparison, we compare scaling on a \textit{per node}
basis. That is, we compare 1 fully occupied 32-core CPU node
(running 32 MPI tasks) with 1 GPU node (host running 1 MPI task,
and 1 device). We believe this is representative of the true ``cost'' of
a simulation in terms of accounting budgets, or electricity.

@@

With these features in mind, we can identify a number of competing
concerns which are relevant to a GPU implementation:
\begin{enumerate}
\item
Minimisation of host-device data transfer would argue for moving the
entire particle code to the GPU. However, the code in question involves
largely conditional logic (e.g., identifying cut surface links) and
irregular memory accesses (e.g., access to distribution elements around
a spherical particle). These operations seems poorly suited to effective
parallelisation on the GPU. As an additional complication, the sums
required over the particle surface would involve potentially tricky
and inefficient reductions in GPU memory.
\item
The alternative is to retain the relevant code on the CPU. While the
transfer of the entire distribution $f_i(\mathbf{r};t)$ between host
and device at each time step is unconscionable owing to memory bandwidth
considerations, the transfer of only relevant distribution information
to allow bounce-back on links is possible. At modest solid volumes
fractions only a very small fraction of the distribution data is
involved in bounce-back (e.g., a 2\% solid volume fraction of particles
radius 2.3$\Delta r$ would involve approximately 1\% of the distribution
data). This option also has the advantage that no further host-device
data transfers are necessary to allow the MPI exchanges required for
particle information.
\end{enumerate}

ALAN PLEASE CHECK:
We have implemented the second option as follows. For each sub-domain,
a list of boundary-cutting links is assembled on the CPU which includes
the identity of the relevant element of the distribution. This list,
together with the particle information required to compute the correct
bounce-back term, are transferred to the GPU. The updates to the relevant
elements of the distribution can then take place on the GPU. The
corresponding information to compute the update of the particle dynamics
is returned
to the CPU, where the reduction over the surface links is computed.
The change of particle shape may be dealt with in a similar manner:
the relatively small number of updates required at any one time step
(or however frequently the particle position is updated) can be
marshaled to the GPU as necessary.
CAN WE SAY SOMETHING LIKE: This preliminary implementation
is found to be effective on problems involving up to X per cent solid
volume fraction.

>>>>

In the CPU version, the \textit{Ludwig} implementation stores one time
level of distribution values $f_i(\mathbf{r}; t)$. This distribution
is stored as a flat array of C data type \texttt{double}, and laid out
so that all elements of the velocity are contiguous at a given site
(often referred to as ``array-of-structures'' order). This is
appropriate for sums over the distribution required locally at the
collision stage as illustrated schematically in
Listing~\ref{ch14:listing1}: the fact that consecutive loads are from
consecutive memory addresses allows the prefetcher to fully engage.
(The temporary scalar \texttt{a\_tmp} allows caching of the
intermediate accumulated value in the innermost loop.)  A variety of
standard sequential optimisations are relevant for the collision stage
(loop unrolling, inclusion of SIMD operations, and so on
\cite{wellein2006}).  For example, the collision stage in
\textit{Ludwig} includes explicit SIMD code, which is useful if the
compiler on a given platform cannot identify it.  The propagation
stage is separate, and is organised as a ``pull'' (again, various
optimisation have been considered, e.g.,
\cite{pohl2003,mattila2007,wittmann2012}).  No further optimisation is
done here beyond ensuring that the ordering of the discrete velocities
allows memory access to be as efficient as possible. While these
optimisations are important, it should be remembered that for some
complex fluid problems, the hydrodynamics embodied in the LB
calculation is a relatively small part of the total computational cost
(at the 10\%-level in some cases).  This means optimisation effort may
be better concentrated elsewhere.

@@

\begin{figure}[!t]
\centering
\includegraphics[width=12cm]{Chapters/chapter14/figures/decomphalo}
\caption{Left: the lattice is decomposed between MPI tasks. For
  clarity we show a 2D decomposition of a 3D lattice, but in practice
  we decompose in all 3 dimensions. Halo cells are added to each
  sub-domain (as shown on the upper right for a single slice) which store
  data retrieved from remote neighbours in the halo exchange. Lower
  right: the D3Q19 velocity set resident on a lattice site;
  highlighted are the 5 ``outgoing'' elements to be transferred in a
  specific direction.}
\label{ch14:fig:decomphalo}
\end{figure}

@@

In this section we describe the steps taken to enable \textit{Ludwig}
for the GPU. There are a number of crucial issues: first, the
minimisation of data traffic between host and device; second, the
optimal mapping of available parallelism onto the architecure and
third, the issue of memory coalescing. We discuss each of these in
turn.

While the most important section of the LB in terms of
floating-point performance is the collision stage, this cannot be the
only consideration for a GPU implementation. It is essential to
offload all computational activity which involves the main data
structures (such as the distribution) to the GPU. This includes
kernels with relatively low computational demand, such as the
propagation stage. All relevant data then remain resident on the GPU,
to avoid expensive host-device data transfer at each iteration of
the algorithm. Such transfers would negate any benefit of GPU
acceleration.  We note that for a complex fluid code, this requirement
can extend to a considerable number of kernels, although we limit the
discussion to collision and propagation for brevity.

@@

To demonstrate the effectiveness of our approach, we compare the
performance of both CPU and GPU versions of \textit{Ludwig}. To
test the complex fluid nature of the code, the problem is
actually an immiscible fluid mixture which uses a second distribution
function to introduce a composition variable. The interested reader
is referred to \cite{ch14:stratford-jsp2005} for further details. The
largest total problem size used is $2548\times 1764\times 1568$.
The CPU system features a Cray XE6 architecture with 2 16-core AMD Opteron
CPUs per node, and with nodes interconnected using Cray Gemini technology.
For GPU results, a Cray XK6 system is used: this is very similar to the
XE6, but has one CPU per node replaced with an NVIDIA X2090 GPU. Each
node in the GPU system therefore features a single Opteron CPU acting
as a host to a single GPU. The inter-node interconnect architecture is
the same as for the Cray XE6. The GPU performance tests use a prototype
Cray XK6 system with 936 nodes (the largest available at the time of writing).
To provide a fair comparison, we compare scaling on a \textit{per node}
basis. That is, we compare 1 fully occupied 32-core CPU node
(running 32 MPI tasks) with 1 GPU node (host running 1 MPI task,
and 1 device). We believe this is representative of the true ``cost'' of
a simulation in terms of accounting budgets, or electricity.

@@

With these features in mind, we can identify a number of competing
concerns which are relevant to a GPU implementation:
\begin{enumerate}
\item
Minimisation of host-device data transfer would argue for moving the
entire particle code to the GPU. However, the code in question involves
largely conditional logic (e.g., identifying cut surface links) and
irregular memory accesses (e.g., access to distribution elements around
a spherical particle). These operations seem poorly suited to effective
parallelisation on the GPU. As an additional complication, the sums
required over the particle surface would involve potentially tricky
and inefficient reductions in GPU memory.
\item
The alternative is to retain the relevant code on the CPU. While the
transfer of the entire distribution $f_i(\mathbf{r};t)$ between host
and device at each time step is unconscionable owing to PCIe bus bandwidth
considerations, the transfer of only relevant distribution information
to allow bounce-back on links is possible. At modest solid volumes
only a very small fraction of the distribution data is
involved in bounce-back (e.g., a 2\% solid volume fraction of particles
radius 2.3$\Delta r$ would involve approximately 1\% of the distribution
data). This option also has the advantage that no further host-device
data transfers are necessary to allow the MPI exchanges required for
particle information.
\end{enumerate}

We have implemented the second option as follows. For each sub-domain,
a list of boundary-cutting links is assembled on the CPU which includes
the identity of the relevant element of the distribution. This list,
together with the particle information required to compute the correct
bounce-back term, are transferred to the GPU. The updates to the relevant
elements of the distribution can then take place on the GPU. The
corresponding information to compute the update of the particle dynamics
is returned
to the CPU, where the reduction over the surface links is computed.
The change of particle shape may be dealt with in a similar manner:
the relatively small number of updates required at any one time step
(or however frequently the particle position is updated) can be
marshaled to the GPU as necessary.
CAN WE SAY SOMETHING LIKE: This preliminary implementation
is found to be effective on problems involving up to X per cent solid
volume fraction.
